---
title: "Reconstruction of EOD Waveform Type Evolution in Paramormyrops kingsleyae using Parsimony, Stochastic Character Mapping and Joint Ancesteral Reconstruction"
output: html_notebook
params:
  data_path: NULL
  output_file: NULL
---
```{r setup, include=FALSE}
root <- rprojroot::find_root(".git/index")
knitr::opts_knit$set(root.dir = root)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(crop = TRUE)
library(tidyverse)  # Includes ggplot2 and other data manipulation packages
library(data.table)
library(ape)
library(phytools)
library(castor)
library(corHMM)
```

# Set up analysis
Specify the paths for the various input phylogenetic trees and phenotypic data
```{r}
treefile<-file.path(root,"output_data/04_Phylogeny_And_Admixture/MIC4273_HAP2_NEWREF_MSU618_all_fish_all_sites.FILTERED_BIALLELIC_SNPS_ONLY.renamed_extracted_gcf_tree.cf.tree")
fish_data_file<-file.path(root,"input_data/01_Terra/data_model/fish_data_2024a.txt")
```

```{r}
snp_tree<-read.tree(treefile)
fish_data<-read.csv(fish_data_file,row.names=1,sep="\t")

pheno_discrete <- fish_data %>% 
  mutate(Phenotype = factor(Phenotype)) %>%
  select(Phenotype)
```

Is there any missing samples between the phenotype file and the tree?
```{r}
phylo_samps<-snp_tree$tip.label
pheno_discrete_samps<-rownames(pheno_discrete)
setdiff(phylo_samps,pheno_discrete_samps)
```

# Clean Up Tree

For reasons that I can't really understand, joint reconstruction fails because of the long branches of PVAD_5505 AND PSZA_454.  So here, I'm going to drop PVAD_5505 and root the tree in PSZA_454.
```{r}
rooted_snp_tree<-root(snp_tree,outgroup="PSZA_454", resolve.root=TRUE)

rooted_snp_tree<-drop.tip(rooted_snp_tree,"PVAD_5505")
```

# Create a Time Calibrated Chronogram
Estimates of crown age come from timetree.org and Peterson et al. 2023 phylogeny
```{r}
mycalibration=makeChronosCalib(rooted_snp_tree,node="root",age.min=8.30,age.max=9.96)
ultrametric_rooted_tree <- chronos(rooted_snp_tree, lambda = .1, model = "relaxed", calibration = mycalibration, control = chronos.control(dual.iter.max=100) )
```

# Organize Phenotypes
Need to categorize assigned phenotypes as TP and BP, and for now we are converting wobbles for illustrative purposes.
```{r}
phenos<-setNames(as.character(pheno_discrete$Phenotype), row.names(pheno_discrete))
phenos <- gsub("TYPE_II", "TP", phenos)
phenos <- gsub("TYPE_I", "TP", phenos)
phenos<-as.factor(phenos)
## set colors for plotting 

newphenos<-phenos[snp_tree$tip.label]
newphenos['APA_193']<-'TP'
newphenos['BAM_165']<-'BP'
newphenos<-droplevels(newphenos)
cols<-setNames(c("blue","red"),levels(newphenos)) 
```

# View Extant Character Distribution on the Phylogeny
```{r}
plotTree(ultrametric_rooted_tree
  ,mar=c(4.1,1.1,0.1,1.1))
axis(1)
title(xlab="millions of years since mrca")
abline(v=seq(0,10,by=2),lty="dotted",col="grey")
tiplabels(pie=to.matrix(newphenos[ultrametric_rooted_tree$tip.label], levels(newphenos)), piecol=cols, cex=0.3)
```
# Using Parsimony to reconstruct character evolution.
See pg. 251 of Revell & Harmon (2022) "Comparative Phylogenetic Comparative Methods in R" - assumes rate of evolution is very slow.
```{r,fig.width=12}
gp.int<-as.numeric(newphenos);
names(gp.int)<-names(newphenos)  
gp.int <- gp.int[match(ultrametric_rooted_tree$tip.label,names(gp.int))]
anc.mp <- asr_max_parsimony(ultrametric_rooted_tree,gp.int)
plot(ultrametric_rooted_tree)
tiplabels(pie=model.matrix(~as.factor(gp.int)-1),piecol=cols,cex=0.3)
nodelabels(node=1:ultrametric_rooted_tree$Nnode+Ntip(ultrametric_rooted_tree),
           pie=anc.mp$ancestral_likelihoods,piecol=cols,cex=0.3)
```
# Joint Character Estimation
See pg. 238 Revell & Harmon (2022) "Comparative Phylogenetic Comparative Methods in R":
In joint ancestral state reconstruction, what we do is ask which set of character values at all the internal nodes maximizes the probability of obtaining the values for the character that we’ve observed at the tips of the tree, given our model.

Approach outlined here: http://blog.phytools.org/2024/04/a-few-useful-demos-on-ancestral-state.html
```{r}
## fit ER model 
fitER<-fitMk(ultrametric_rooted_tree,newphenos,model="ER", pi=c(0,1)) 
## fit ARD model 
fitARD<-fitMk(ultrametric_rooted_tree,newphenos,model="ARD",pi=c(0,1))
## fit SYM model
fitSYM<-fitMk(ultrametric_rooted_tree,newphenos,model="SYM",pi=c(0,1))
```

```{r}
model_aov<-anova(fitER,
  fitARD,fitSYM)

```


```{r}
paramormyrops_ancr<-ancr(fitER, type="joint")
paramormyrops_ancr
```
```{r, fig.width=12}
obj<-plot(paramormyrops_ancr,
  legend=FALSE,piecol=cols)
legend("bottomright",levels(newphenos),
  pch=16,col=obj$piecol,bty="n",cex=0.8,
  pt.cex=1.5)

```
```{r}
wavetype_hrm<-fitHRM(ultrametric_rooted_tree,newphenos,ncat=2,pi=(c(0,1)),parallel=TRUE,niter=20)

```