---
title: "R Notebook"
output: html_notebook
---

#Analyses for Tutorial 06: Ancestral State Estimation in R
  #Parts of this tutorial are based on that of L. Revell: http://www.phytools.org/eqg2015/asr.html

```{r setup, include=FALSE}

root <- rprojroot::find_root(".git/index")
knitr::opts_knit$set(root.dir = root)

library(tidyverse)  # Includes ggplot2 and other data manipulation packages

library(phytools)
library(geiger)
library(castor)
library(corHMM)
```


```{r}
treefile<-file.path(root,"output_data/04_Phylogeny_And_Admixture/MIC4273_HAP2_NEWREF_MSU618_all_fish_all_sites.FILTERED_BIALLELIC_SNPS_ONLY.renamed_extracted_gcf_tree.cf.tree")
fish_data_file<-file.path(root,"input_data/01_Terra/data_model/fish_data_2024a.txt")

snp_tree<-read.tree(treefile)
fish_data<-read.csv(fish_data_file,row.names=1,sep="\t")

pheno_discrete <- fish_data %>% 
  mutate(Phenotype = factor(Phenotype)) %>%
  select(Phenotype)

phenos<-setNames(as.character(pheno_discrete$Phenotype), row.names(pheno_discrete))
phenos <- gsub("TYPE_II", "TP", phenos)
phenos <- gsub("TYPE_I", "TP", phenos)
phenos<-as.factor(phenos)

newphenos<-phenos[snp_tree$tip.label]
newphenos['APA_193']<-'TP'
newphenos['BAM_165']<-'BP'
newphenos<-droplevels(newphenos)
cols<-setNames(c("blue","red"),levels(newphenos)) 


rooted_snp_tree<-root(snp_tree,outgroup=c("PSZA_454"), resolve.root=TRUE)

rooted_snp_tree$node.label<-NULL

```

```{r, fig.height=8}
#Plot 
plot(rooted_snp_tree)
tiplabels(pie=to.matrix(phenos[rooted_snp_tree$tip.label], levels(phenos)), piecol=cols, cex=0.3)
```
```{r,fig.height=8}
gp.int<-as.numeric(newphenos); names(gp.int)<-names(newphenos)  
anc.mp <- asr_max_parsimony(rooted_snp_tree,gp.int)

plot(rooted_snp_tree)
tiplabels(pie=model.matrix(~as.factor(gp.int)-1),piecol=cols,cex=0.3)
nodelabels(node=1:rooted_snp_tree$Nnode+Ntip(rooted_snp_tree),
           pie=anc.mp$ancestral_likelihoods,piecol=cols,cex=0.3)

```

```{r}
##Create the model for trait evolution
group_formatted<-cbind(names(newphenos),as.character(newphenos)) ##put the data in a specific format because the function is picky
group_model_er <- getStateMat4Dat(group_formatted,"ARD")  #equal transition rates
plotMKmodel(group_model_er$rate.mat,rate.cat = 1)
```

```{r}
##Perform ancestral character estimation
anc.ML<-ace(x = newphenos, phy = rooted_snp_tree, 
            type = "discrete", method = 'ML',
            model = group_model_er$rate.mat)

anc.ML ## we can see a lot of information just by calling our object

## We can see the probability of each state here, just the first few internal nodes will be shown
round(head(anc.ML$lik.anc),3)

```

```{r,fig.height=8}}
plot(rooted_snp_tree)
nodelabels(pie=anc.ML$lik.anc,piecol=cols,cex=0.5)
tiplabels(pch=19,col=cols[newphenos])
legend(x='bottomleft',legend = names(cols),fill=cols)

```


## ---- simmap ----
# simulate single stochastic character map using empirical Bayes method
gp<-setNames(group$x,rownames(group)) ##make.simmap needs a named vector for the data.
tree.smp1<-make.simmap(tree,gp,model=group_model_er$rate.mat)
plot(tree.smp1,colorkey,type="fan")  #one run.  Not overly useful. 

#Must do many times
tree.smp<-make.simmap(tree,gp,model=group_model_er$rate.mat,nsim=100)
anc.smp<-summary(tree.smp,plot=FALSE)
plot(anc.smp, type="fan", fsize=0.8,ftype="i")

rm(list = ls())

## ---- read_data_cont ----

#Data from Mahler et al. 2010. Evolution
tree<-read.tree("../data/anole.svl.tre",tree.names=T)
cont_data<-read.csv('../data/anole.svl.csv', row.names=1)
cont_data_vect<-setNames(cont_data$svl,rownames(cont_data)) ##contMap needs a named vector for the data.

## -----anc_est_cont-----
anc.cont.ML<-fastAnc(tree,cont_data_vect,vars=TRUE,CI=TRUE)
#anc.cont.ML #This would show us the estimate and 95% CI for each internal node
anc.cont.ML$ace  #ancestral estimates

#PLOT as color map
tree.col<-contMap(tree,cont_data_vect,plot=FALSE)  #runs Anc. St. Est. on branches of tree
plot(tree.col,type="fan")

## ----anc_est_cont_ace_ml----

#using 'ace' function in APE
anc.cont.ML2<-ace(x=cont_data_vect,phy=tree, type="continuous", method="ML")
anc.cont.ML2$ace  #with APE
anc.cont.ML$ace   #with phytools: the same estimates


## ----anc_est_cont_ace_gls ----

anc.cont.gls<-ace(x=cont_data_vect,phy=tree, corStruct = corBrownian(1, tree), method="GLS")  #same as ML  (see Schluter et al. 1997)
anc.cont.gls$ace   #GLS: SCP. the same

rm(list = ls())


## ---- known_nodes ----
tree<-pbtree(n=100,scale=1)
## simulate data with a trend
x<-fastBM(tree,internal=TRUE,mu=3)
phenogram(tree,x,ftype="off")

x.tip<-x[match(tree$tip.label,names(x))]
phenogram(tree,x.tip)  #traitgram under BM with no ancestral information

#estimate with no prior 
a<-x[as.character(1:tree$Nnode+Ntip(tree))]
x<-x[tree$tip.label]
## let's see how bad we do if we ignore the trend
plot(a,fastAnc(tree,x),xlab="true values",
     ylab="estimated states under BM")
lines(range(c(x,a)),range(c(x,a)),lty="dashed",col="red")
title("estimated without prior information")

## incorporate prior knowledge
pm<-setNames(c(1000,rep(0,tree$Nnode)),
             c("sig2",1:tree$Nnode+length(tree$tip.label)))
## the root & two randomly chosen nodes
nn<-as.character(c(length(tree$tip.label)+1,
                   sample(2:tree$Nnode+length(tree$tip.label),2)))
pm[nn]<-a[as.character(nn)]
## prior variance
pv<-setNames(c(1000^2,rep(1000,length(pm)-1)),names(pm))
pv[as.character(nn)]<-1e-100
## run MCMC
mcmc<-anc.Bayes(tree,x,ngen=100000,
                control=list(pr.mean=pm,pr.var=pv,
                             a=pm[as.character(length(tree$tip.label)+1)],
                             y=pm[as.character(2:tree$Nnode+length(tree$tip.label))]))

anc.est<-colMeans(mcmc$mcmc[201:1001,as.character(1:tree$Nnode+length(tree$tip.label))])

plot(a,anc.est,xlab="true values",
     ylab="estimated states using informative prior")
lines(range(c(x,a)),range(c(x,a)),lty="dashed",col="red")
title("estimated using informative prior")
