---
title: "Phylogenetic Reconstruction of BP, TP and Wobbles"
output: html_notebook
params:
  data_path: NULL
  output_file: NULL
---
```{r setup, include=FALSE}
root <- rprojroot::find_root(".git/index")
knitr::opts_knit$set(root.dir = root)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(crop = TRUE)
library(tidyverse)  # Includes ggplot2 and other data manipulation packages
library(data.table)
library(ape)
library(phytools)
library(castor)
```

```{r}
snp_tree<-paramormyropstree
fish_data<-read.csv2('~/Desktop/paramormyrops_phenotypes.csv',row.names=1,header=F,sep=",")
```

Is there any missing samples between the phenotype file and the tree?
```{r}
phylo_samps<-snp_tree$tip.label
pheno_discrete_samps<-rownames(fish_data)
setdiff(phylo_samps,pheno_discrete_samps)
```

# Organize phenotypes so that there are three (TP, BP and Wobble)
```{r}
phenos<-setNames(as.character(fish_data$V2), row.names(fish_data))
phenos<-as.factor(phenos)
## set colors for plotting 
cols<-setNames(c("blue","red"),levels(phenos)) 

```

```{r}
plotTree(snp_tree
  ,mar=c(4.1,1.1,0.1,1.1))
axis(1)
title(xlab="relative time before present")
abline(v=seq(0,15,by=5),lty="dotted",col="grey")
tiplabels(pie=to.matrix(phenos[snp_tree$tip.label], levels(phenos)), piecol=cols, cex=0.3)
```



```{r}
## fit ER model 
fitER<-fitMk(snp_tree,phenos,model="ER") 
## fit ARD model 
fitARD<-fitMk(snp_tree,phenos,model="ARD") 
## fit SYM model
fitSYM<-fitMk(snp_tree,phenos,model="SYM")
```

```{r}
## extract AIC values for each model 
aic<-c(AIC(fitER),AIC(fitARD),AIC(fitSYM)) 
## print summary table 

model_summary<-data.frame(
  model=c("ER","ARD","Symmetric"), 
  logL=c(logLik(fitER),logLik(fitARD),logLik(fitSYM)), 
  AIC=aic,delta.AIC=aic-min(aic),
  weight=unclass(aic.w(aic)))

model_summary
```
```{r, fig.width=4, fig.height=4}
  plot(fitER, width=TRUE, color=TRUE, cex.rates=0.9, main="ER Model") # Add a main title with the tree name
```

## Perform Stochastic Character Mapping

See Revell & Harmon (2022) "Comparative Phylogenetic Comparative Methods in R", page 245.

```{r}
mtrees<-make.simmap(snp_tree,phenos,model="ER",nsim=1000,Q="mcmc",vQ=0.01, prior=list(use.empirical=TRUE),samplefreq=10)
```
```{r}
pd<-summary(mtrees)
pd
```

Let's examine the distribution of changes that we see.

Interestingly, about 6 transitions directly from TP->BP (on average) (see http://blog.phytools.org/2022/07/understanding-number-of-changes-of.html)
```{r}
dd<-density(mtrees)
dd
```

We can look at these densities graphically as well...
```{r, fig.height=12}
plot(dd,colors=c("orange","lightgreen","lightblue"),alpha=0.5)

```
```{r}
all_trees<-as.data.frame(countSimmap(mtrees,c("BP","TP")))

all_trees<-rownames_to_column(all_trees, "treeno")
all_trees<-all_trees %>% arrange(Tr.N)
head(all_trees)
```
```{r, fig.width=12}
## create a plot showing PP at all nodes of the tree 
plot(pd,colors=cols,fsize=1,ftype="i",lwd=2,
offset=0.4,ylim=c(-1,Ntip(snp_tree)),cex=c(0.5,0.3)) 
## add a legend
legend("bottomleft",legend=levels(phenos),pch=22, pt.cex=1.5,pt.bg=cols,bty="n",cex=0.8)
```

```{r,fig.width=12}
gp.int<-as.numeric(phenos);
names(gp.int)<-names(phenos)  
gp.int <- gp.int[match(snp_tree$tip.label,names(gp.int))]
anc.mp <- asr_max_parsimony(snp_tree,gp.int)
plot(snp_tree)
tiplabels(pie=model.matrix(~as.factor(gp.int)-1),piecol=cols,cex=0.3)
nodelabels(node=1:snp_tree$Nnode+Ntip(snp_tree),
           pie=anc.mp$ancestral_likelihoods,piecol=cols,cex=0.3)
```